Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections

### 2023-04-25 paper.qmd
done??

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
subtitle: "Final Project for STAT 530"
author: "Tyler Bagwell and Prayag Gordy"
date: "April 24, 2023"
abstract: "The incumbency effect in U.S. elections is a subject of decades of political science research. Why is it, political scientists ask, that legislative and executive bodies with such low overall approval ratings have the vast majority of its members returning term after term? In this paper, we will use a regression discontinuity design to examine the effects of party incumbency in tightly contested congressional districts. Specifically, we examine the effect of the Democratic margin in an election on the Democratic margin in the following election. We found a local average treatment effect of 17.3 percentage points. Our results further support the literature claiming a large incumbency effect, though we show its strength may be decreasing."
format: pdf
toc: true
number-sections: true
execute:
  echo: false
  message: false
  warning: false
---

@@

lag_any_candidates(DF) |>
  dplyr::pull(prev_dem_margin) |>
  rdd::DCdensity()
```

@@

>>>>

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
subtitle: "Final Project for STAT 530"
author: "Tyler Bagwell and Prayag Gordy"
date: "April 24, 2023"
abstract: "The incumbency effect in U.S. elections is a subject of decades of political science research. Why is it, political scientists ask, that legislative and executive bodies with such low overall approval ratings have the vast majority of its members returning term after term? In this paper, we will use a regression discontinuity design to examine the effects of party incumbency in tightly contested congressional districts. Specifically, we examine the effect of the Democratic margin in an election on the Democratic margin in the following election. We found a local average treatment effect of 17.3 percentage points. Our results further support the literature claiming a large incumbency effect, though we show its strength may be decreasing."
format: pdf
toc: true
number-sections: true
execute:
  echo: false
  message: false
  warning: false
  results: false
---

@@

invisible(rdd::DCdensity(lag_any_candidates(DF)$prev_dem_margin))
```

@@

### Basic RDD Results

With diagnostics assessed, we ran our first RDD analyses. After applying our filters, we had 4,800 observations. We used the `rdd` package in R to conduct our analyses.

We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory. A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$. This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.

```{r}
#| fig-cap: RDD analysis of 4,800 elections from 1976 through 2020. The local average treatment effect is 0.173.

lag_any_candidates(DF) |>
  plot_rdd()
```

We also conducted logistic regressions as part of our analyses. Instead of examining the Democratic margin in election $t$, we consider only the final result: a Democratic victory or loss. These logistic regressions support the narrative from the RDD analysis, but in a simpler setting.

```{r}
#| fig-cap: Separate logistic regressions on either side of the cutoff also show a clear discontinuity, with a 0.551 change in the probability of a Democrat winning election $t$.

logistic(DF)
```

To further separate potential associative relationships from causal relationships, we reversed our analyses. Clearly, an election in year $t$ cannot have an effect on an election in year $t-1$ because of the temporal structure in the data. Our empirical results align with this theory, as there is practically no discontinuity in the RDD or logistic analyses and thus no evidence of retro-causality.

```{r}
#| fig-cap: An RDD applied in reverse — the effect of the Democratic margin in election $t$ on election $t-1$ — shows no discontinuities.

backwards_rdd(DF)
```

```{r}
#| fig-cap: A logistic regression applied in reverse also shows no discontinuities.

logistic_backwards(DF)
```

### Discarded Analyses

Up to this point, our analyses have examined a slightly weaker form of incumbency: party incumbency. That is, we looked at the Democratic margin without considering the candidates themselves. One Democrat may run in election $t-1$ but another in election $t$, and we would treat this identically to a situation in which the same Democrat runs in elections $t-1$ and $t$. We sought to narrow our analysis to eight cases:

- the same Democrat runs in elections $t-1$ and $t$, regardless of the best-performing non-Democrat;
- a different Democrat runs in elections $t-1$ and $t$, regardless of the non-Democrat;
- the same Democrat runs in elections $t-1$ and $t$ against a different non-Democrat;
- the same Democrat runs in elections $t-1$ and $t$ against the same non-Democrat;
- the same non-Democrat runs in elections $t-1$ and $t$, regardless of the Democrat;
- a different non-Democrat runs in elections $t-1$ and $t$, regardless of the Democrat;
- the same non-Democrat runs in elections $t-1$ and $t$ against a different Democrat; and
- the same non-Democrat runs in elections $t-1$ and $t$ against the same Democrat.

However, these sub-groups fail to meet the assumptions required for RDD. This may be because of who runs again. If a Democratic candidate loses by a slim margin, they are far more likely to step away from politics than a Democrat who barely wins, even though the candidates and their districts may be statistically similar. We empirically do not see continuity, which has a theoretical justification.

```{r}
#| fig-cap: Density plots for the first four sub-groups show violations in the continuity assumption.
#| fig-height: 6

par(mfrow = c(2, 2))
invisible(rdd::DCdensity(lag_same_dem(DF)$prev_dem_margin) )
invisible(rdd::DCdensity(lag_diff_dem(DF)$prev_dem_margin))
invisible(rdd::DCdensity(lag_same_dem_diff_rep(DF)$prev_dem_margin) )
invisible(rdd::DCdensity(lag_same_dem_same_rep(DF)$prev_dem_margin))
```

### Additional Results

We did manage to successfully divide our population into some relevant sub groups. First, we divided the elections by decade. As the table shows, the effect of a Democrat winning a close election on the Democratic margin in the next election has decreased over time.

```{r}
#| tbl-cap: The LATE has decreased over time.

decade_rdd(DF) |>
  knitr::kable()
```

Then, we considered the region of the congressional district, as defined by the U.S. Census Bureau. Many regions have similar LATE estimates. The Pacific region, comprising Alaska, Washington, Oregon, California, and Hawaii, has the lowest estimate. The West South Central, containing Oklahoma, Texas, Arkansas, and Louisiana, has the second highest LATE estimate, while New England — Maine, New Hampshire, Vermont, Massachusetts, Connecticut, and Rhode Island — has the highest at 0.269.

```{r}
#| tbl-cap: Most regions see similar LATE estimates, though New England may have a particularly strong incumbency effect.

region_rdd(DF) |>
  knitr::kable()
```

## Conclusions

Our analyses support previous literature showing a strong incumbency effect. Districts that are otherwise similar, as our covariate analysis showed, may have vastly different electoral futures come down to just a handful of votes. These results show the importance of national parties investing in close congressional elections, as a marginal amount of votes in one election can affect following elections.

There are many analyses researchers can conduct to further the study of incumbency advantages. Demographic data on the candidates themselves can be considered to explore racial and other sorts of discrimination in politics. Scientists can examine when a losing candidate runs again: Do they take a few cycles to recover from a grueling and heartbreaking campaign? Do they never re-enter politics? Does it depend on how close they were? Finally, scientists can reproduce these analyses at other levels — the U.S. Senate, state senates, state houses, and local offices.
